---
title: "In-Class Exercise 6: Spatial Weights - sfdep methods"
date: "13 February 2023"
date-modified: "`r Sys.Date()`"
number-sections: true
format: html
execute: 
  eval: true
  echo: true
  warning: false
editor: visual
---

# Installing and Loading R Packages

```{r}
pacman::p_load(sf, tmap, sfdep, tidyverse)
```

# Data

## Importing Geospatial Data

```{r}
hunan <- st_read(dsn = "data/geospatial",
                 layer = "Hunan")
```
## Importing Aspatial Data 

Importing Hunan_2012.csv into a tibble data frame

```{r}
hunan_2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

## Combining both data frame by using left join

In order to retain the geospatial property, the left data frame must be the sf data frame.

```{r}
hunan_GDPPC <- left_join(hunan, hunan_2012) %>%
  select(1:4, 7, 15)

# normally we have to specify which common field to do the join
# R is case sensitive, we need to check if the cases match (or else they will not join)
# After joining, we retain columns 1-4, 7 and 15 of the joined dataframe only
```

## Plotting a choropleth map

```{r}
tmap_mode("plot")
tm_shape(hunan_GDPPC)+
  tm_fill("GDPPC",
          style="quantile",
          palette = "Blues",
          title = "GDPPC") +
  tm_layout(main.title = "Distribution of GDP per capita by district, Hunan Province",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45,
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha = 0.2)
```
# Identify Area Neighbours

## Contiguity Neighbours Method

st_contiguity() derives a contiguity neighbour list using Queen's method (queen method by default, do not need to specify)

```{r}
cn_queen <- hunan_GDPPC %>%
  mutate(nb = st_contiguity(geometry),
         .before = 1)

# mutate creates a new field called nb which is the contiguity neighbour list
# .before = 1 will create the new field as the first column
```

Using Rook's Method:

```{r}
cn_rook <- hunan_GDPPC %>%
  mutate(nb = st_contiguity(geometry),
         queen = FALSE,
         .before = 1)
```

Note: Bishop method does not exist for sfdep (only in spdep)

# Computing contiguity weights

## Queen's Method

```{r}
wm_q <- hunan_GDPPC %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(nb),
         .before = 1)
# can combine the steps of creating the contiguity neighbour list together with computing the contiguity weights

# In the sfdep package
```

## Rook's Method

```{r}
wm_r <- hunan_GDPPC %>%
  mutate(nb = st_contiguity(geometry),
         queen = FALSE,
         wt = st_weights(nb),
         .before = 1)
```
